---
title: "R Coursework 3"
author: '36306330'
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## R Markdown

## Task1 : print dimension

```{r task1}
Aus_storm <- read.csv(file = 'Australia_severe_storms_1975-2015.csv')
dim(Aus_storm)
```

## Task 2: clean the data

```{r task2}
# remove ID & waterspout in Database
Aus_storm <- Aus_storm %>% select(-ID) %>% 
  filter(Database != 'Waterspout')
# print dimension
dim(Aus_storm)
# print first few rows without the 7 columns of comments
head(select(Aus_storm, -Comments, -X, -X.1, -X.2, -X.3, -X.4, -X.5))
```

## Task 3: Add a column containing time zone of each event

```{r task 3}
#if_else to check assign timezone
assign_timezone <- function(state, town){
  if (state == 'QLD'){
    timezone = 'Australia/Queensland'
  }
  if (state == 'NSW'){
    if (str_detect(town, '([Bb][Rr][Oo][Kk][Ee][Nn] [Hh][Ii][Ll][Ll])')){
        timezone = 'Australia/Broken_Hill'
    }
    else {
      timezone = 'Australia/NSW'
    }
  }
  if (state == 'VIC'){
    timezone = 'Australia/Victoria'
  }
  if (state == 'SA'){
    timezone = 'Australia/South'
  }
  if (state == 'WA'){
    timezone = 'Australia/West'
  }
  if (state == 'TAS'){
    timezone = 'Australia/Tasmania'
  }
  if (state == 'NT'){
    timezone = 'Australia/North'
  }
  if (state == 'ACT'){
    timezone = 'Australia/ACT'
  }
  return(timezone)
}
  
#for loop for dataframe
for (i in 1:nrow(Aus_storm)){
  Aus_storm$Timezone[i] <- assign_timezone(Aus_storm$State[i],Aus_storm$Nearest.town[i])
}
```

## Task 4: Converts the time into UTC

```{r task 4}
#for loop to convert the data to date time format according timezone
results <- POSIXct()
for (i in 1:nrow(Aus_storm)){
  results[i] <- dmy_hm(Aus_storm$Date.Time[i], tz = Aus_storm$Timezone[i])
}
Aus_storm$Date.Time.UTC <- results

#Print the first few rows without the 7 columns of comments
head(select(Aus_storm, -Comments, -X, -X.1, -X.2, -X.3, -X.4, -X.5))
```

## Task 5: Create new variables for the month and year of each event

```{r task 5}
#create variable month & year
Aus_storm$Date.Time <- dmy_hm(Aus_storm$Date.Time)
Aus_storm$Month <- month(Aus_storm$Date.Time)
Aus_storm$Year <- year(Aus_storm$Date.Time)

# #Print the first few rows without the 7 columns of comments
head(select(Aus_storm, -Comments, -X, -X.1, -X.2, -X.3, -X.4, -X.5))
```

## Task 6:

i)  Create a new data frame: total number of counts of each events for each twelve months over the forty year period
```{r task 6i}
Events_month <- Aus_storm %>% count(Month, Event = Database)
```

ii) Plot total number of counts of each event against month
```{r task 6ii}
ggplot(Events_month) +
  geom_line(aes(x = Month, y = n, color = Event)) +
  theme(legend.position = 'top') +
  ylab('Total counts') +
  xlab('Months') +
  scale_x_discrete(limits = month.abb)
```

## Task 7: Modify Comments columns

i)  Combine the comments into a single column: All.comments
```{r task 7i}
Aus_storm <- Aus_storm %>% unite('All.comments', Comments:X.5, sep = " ")
```

ii) Select Event.ID, Database, State, All.comments, and the year variable created.
```{r task 7ii}
Aus_storm <- Aus_storm %>% select(Event.ID, Database, State, All.comments, Year)
```

iii) print(sapply(DF, class))
```{r task 7iii}
print(sapply(Aus_storm, class))
```

## Task 8:

i)  Create an indicator variable which states whether or not a storm event has resulted in a flash flood.
```{r task 8i}
expr <- '([Ff]lash\\s*[Ff]lood\\w*)'
Aus_storm$flash_flood <- str_detect(Aus_storm$All.comments, expr)
```

ii) Plot of the number of flash floods per year from 1975-2015. [3 marks]
```{r task 8ii}
#create a data frame to store number of flash floods and year
#filter true only for flash flood
flash_flood_by_year <- filter(Aus_storm, flash_flood == T)
flash_flood_by_year <- count(flash_flood_by_year, Year, flash_flood)

ggplot(flash_flood_by_year) +
  geom_line(aes(x = Year, y = n)) + 
  xlab('Year') +
  ylab('Number of flash flood') +
  ggtitle('Number of flash flood from 1975 to 2015')
```

## Task 9:

i)  Extract all wind speeds (knots and km/h)
```{r}
# km/h, km/hr, Km/hr, km.hr
# kmph, kmh, kph,
# kt, kts, KT 
# KNOTS, knot, knots 
# kn 

expr <- '\\b\\d{1,3}\\s*[Kk][Tt][sS]?|\\d+\\s*kn|\\d+\\s*[Kk][Nn][Oo][Tt][Ss]?
|\\d{1,3}\\s*[kK]m[./]hr?|\\d+\\s*k[mp]p?h'
Aus_storm$wind.speed <- str_extract(Aus_storm$All.comments, expr)

# \\b\\d{1,3} to capture 1 to 3 digits of wind speed, some of the datas
# have extremely large values (not possible for wind speed)
```

ii) Convert km/h wind speeds to knots (1 knot = 1.852 km/h) rounding the wind/speed to the nearest knot.
```{r}
#filter dataframe to exclude wind speed NA
Aus_storm <- Aus_storm %>% filter(!is.na(wind.speed))

#extract to a vector
ws_knot <- as.numeric(str_extract(Aus_storm$wind.speed, '\\d+'))

for (i in 1:nrow(Aus_storm)){
if (str_detect(Aus_storm$wind.speed[i], '\\d{1,3}\\s*[kK]m[./]hr?|\\d+\\s*k[mp]p?h'))
  {
    Aus_storm$wind.speed[i] <- round(ws_knot[i] / 1.852, digit = 0)
  }
else {
Aus_storm$wind.speed[i] <- ws_knot[i]
} 
  }

#rename col to specifiy unit knot
Aus_storm <- Aus_storm %>% rename(wind.speed.knot = wind.speed)
```

iii) Boxplot of the wind speeds recorded per state.

```{r}
#change col class to numeric
Aus_storm$wind.speed.knot <- as.numeric(Aus_storm$wind.speed.knot)

#Boxplot
ggplot(Aus_storm) +
  geom_boxplot(aes(x = State, y = wind.speed.knot)) +
  ylab('Wind Speed (knot)') +
  ggtitle('Wind speed per State')
```
